Isomorphic Interpreters from Logically Reversible Abstract Machines

### 2012-05-29 iso-int/rc-iso-int.tex
fixes

% The constructions in this paper apply only to a class of interpreters
% that can be represented by logically reversible small-step operational
% semantics. While, in a formal sense, they are related to Abramsky's
% ``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
% to give them a more direct characterization purely in terms of logical
% reversibility of the small-step semantics.

@@

@@

\vspace{-20px}
\begin{multicols}{2}
\begin{center}
\scalebox{0.5}{
  \includegraphics{diagrams/nat-nat6.pdf}
}
\end{center}
\columnbreak
\begin{center}
\scalebox{0.6}{
  \includegraphics{diagrams/nat-nat7.pdf}
}
\end{center}  
\end{multicols}

@@

% \paragraph*{Conditions that must be satisfied for constructing a isomorphic
% interpreter}:
% \begin{enumerate}
% \item Distinguishable start and stop states. There may be more than one
%   valid start state and more than one valid stop state.
% \item Each valid machine state must match a unique reduction LHS. Each
%   valid machine state must machine a unique reduction RHS.
% \item Every reduction step must be computable (in {{langRevT}}) and
%   must be logically reversible -- i.e. it must be possible to reduce
%   from right to left.
% \end{enumerate}

@@

@@

We can now apply our construction. The first step to developing the
isomorphic interpreter is to recognize that the two possible kinds of machine
states simply hide an implicit {{bool=T~|~F}}. We make this explicit:

\begin{small}
\begin{multicols}{2}
%subcode{bnf} include main
% Machine states = <bool, t, c>
%
% Start state = <T, t, []>
% Stop state =  <F, t, []>

%subcode{opsem} include main
%! columnStyle = rclr
% <T, L n, c> &|-->& <F, L (n+1), c> &~~~~~ rule 1
% <T, N t1 t2, c> &|-->& <T, t1, Lft c t2> & rule 2 
% <F, t1, Lft c t2> &|-->& <T, t2, Rgt t1 c> & rule 3
% <F, t2, Rgt t1 c> &|-->& <F, c, N t1 t2> & rule 4
\end{multicols}
\end{small}

@@

On the input side, for {{T}} states, we have expanded the tree component and
for {{F}}~states we have expanded the tree contexts. Similarly we have done
the opposite on the output side exactly matching up what the abstract machine
does. One thing to note is that we dropped the {{1}} introduced by expanding
the {{bool}} and instead just labeled the {{true}} and {{false}} branches.

We are ready to start connecting the machine states corresponding to
the reductions that we would like:
\begin{enumerate}
\item When we encounter a leaf we would like to increment its value and move
  to the corresponding {{F}} machine state. For the sake of simplicity, in
  this interpreter we won't be concerned with exposing stuck states: we
  simply use {{add1}} whose adjoint {{sub1}} diverges when applied to {{0}}.

>>>>

@@

\paragraph*{Universality.} 
As an example, consider the implementation of the Toffoli gate below,
which establishes that {{langRevT}} is complete for combinational
circuits. This encoding uses the type {{1+1}} as {{bool}} and values
{{left ()}} and {{right()}} as {{true}} and {{false}}. Previous
work\cite{James:2012:IE:2103656.2103667} (see Sec 5) also establishes
that {{langRevT}} is Turing complete by showing the compilation of a
Turing complete language -- a first-order language with numbers and
iteration -- to {{langRevT}}.

\begin{center}
  \includegraphics{diagrams/toffoli.pdf}
\end{center}

@@

\vspace{-20px}
% \begin{multicols}{2}
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/nat-nat6.pdf}
}
\end{center}

\noindent
we exchange sections A and B while maintaining the connectivity of the wires:

%\columnbreak
\begin{center}
\scalebox{1.0}{
  \includegraphics{diagrams/nat-nat7.pdf}
}
\end{center}  
%\end{multicols}

@@

@@

%% In the absence of a formal classification for styles of reversible
%% abstract machines, we will present only an informal characterisation
%% of our input language.

\paragraph*{Input Machines.}

Our translation apply to machines presented as small-step operational
semantics, i.e. machines that are step-wise rewrites of tuples of
algebraic datatypes, wherein each rewrite step must be logically
reversible. The formal criterion logical reversibility in this setting
are captured in Abramsky's presentation of \emph{bi-orthogonal
  automata} \cite{abramsky2005structural} and may be summarized as:
\begin{enumerate}
\item
  Machines must have distinguishable start and stop states. There may
  be more than one valid start state and more than one valid stop
  state.
\item Each valid machine state must match a unique pattern on the
  left-hand-side and right-hand-side of the `{{|-->}}' rewrite.
\item Every reduction step must be computable (in {{langRevT}}) and
  must be logically reversible -- i.e. it must be possible to reduce
  from right to left.
\end{enumerate}

%% The constructions in this paper apply only to a class of interpreters
%% that can be represented by logically reversible small-step operational
%% semantics. While, in a formal sense, they are related to Abramsky's
%% ``bi-orthogonal automata''\cite{abramsky2005structural}, we will try
%% to give them a more direct characterization purely in terms of logical
%% reversibility of the small-step semantics.

\paragraph*{Steps of the Translation.}

@@

We can now apply our construction. The first step to developing the
isomorphic interpreter is to recognize that the two possible kinds of machine
states simply hide an implicit {{bool}}. We make this explicit:

\begin{small}
\begin{multicols}{2}
%subcode{bnf} include main
% Machine states = <bool, t, c>
%
% Start state = <true, t, []>
% Stop state =  <false, t, []>

%subcode{opsem} include main
%! columnStyle = rclr
% <true, L n, c> &|-->& <false, L (n+1), c> &~~~~~ rule 1
% <true, N t1 t2, c> &|-->& <true, t1, Lft c t2> & rule 2 
% <false, t1, Lft c t2> &|-->& <true, t2, Rgt t1 c> & rule 3
% <false, t2, Rgt t1 c> &|-->& <false, c, N t1 t2> & rule 4
\end{multicols}
\end{small}

@@

On the input side, for {{true}} states, we have expanded the tree component and
for {{false}}~states we have expanded the tree contexts. Similarly we have done
the opposite on the output side exactly matching up what the abstract machine
does. One thing to note is that we dropped the {{1}} introduced by expanding
the {{bool}} and instead just labeled the {{true}} and {{false}} branches.

We are ready to start connecting the machine states corresponding to
the reductions that we would like:
\begin{enumerate}
\item When we encounter a leaf we would like to increment its value and move
  to the corresponding {{false}} machine state. For the sake of simplicity, in
  this interpreter we won't be concerned with exposing stuck states: we
  simply use {{add1}} whose adjoint {{sub1}} diverges when applied to {{0}}.
